---
title: "🏆 Simulating FIFA World Cup 2026"
subtitle: "Exploratory analysis, Monte Carlo simulation and data visualization"
author: "Joel Giroud"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    theme: darkly
    highlight: zenburn
    code_folding: show
    number_sections: true
    df_print: paged
    self_contained: true
---

# =============================================================================
# INTRODUCTION
# =============================================================================

This project analyzes the **FIFA World Cup 2026**, the first in history to feature
48 national teams across 3 host countries (Mexico, United States, and Canada).
The goal is to answer 3 questions through data analysis:

1. **How are the matches distributed across venues, groups, and dates?**
2. **Which team has the highest probability of winning the tournament?**
3. **How does Mexico's local hosting advantage impact its probabilities?**

**Methodology:**  
This project uses the January 2026 FIFA rankings as a quantitative proxy for each
team's strength, as they summarize recent performance in a single score — weighting
opponents, match importance, and results — similarly to Elo systems used in the
sports modeling literature.

From those scores, we build a logistic model that translates the points difference
between two teams into a win probability, since logistic regression is a standard
approach for estimating binary event probabilities such as "A wins vs A does not win"
in sports models (see non-deterministic algorithms and binomial distributions).

On top of that single-match model, we apply a Monte Carlo simulation (widely accepted
in tournament modeling): we generate 10,000 complete tournaments, respecting the
group stage and knockout format, to quantify the distribution of possible outcomes
(champion, finalist, etc.) rather than a single point estimate — which is precisely
Monte Carlo's advantage when uncertainty and bracket combinations are numerous.

---

## Environment setup
We load the required packages and define file paths. All datasets are stored
outside the project in `../datasets/` to keep the repository clean.

# =============================================================================
# 01 - Initial Validation & Merge — World Cup 2026
# =============================================================================

```{r setup, include=TRUE}
# ✅ FIX: each if (!require) installs its own package
if (!require("dplyr"))       install.packages("dplyr")
if (!require("readr"))       install.packages("readr")
if (!require("tidyverse"))   install.packages("tidyverse")
if (!require("janitor"))     install.packages("janitor")
if (!require("lubridate"))   install.packages("lubridate")
if (!require("pointblank"))  install.packages("pointblank")
if (!require("kableExtra"))  install.packages("kableExtra")
if (!require("countrycode")) install.packages("countrycode")
if (!require("ggimage"))     install.packages("ggimage")
if (!require("ggtext"))      install.packages("ggtext")

library(dplyr)
library(readr)
library(tidyverse)
library(janitor)
library(lubridate)
library(pointblank)
library(kableExtra)
library(countrycode)
library(ggimage)
library(ggtext)
library(ggplot2)

PATH_RAW      <- "../datasets/FIFA World Cup 2026 Match Data (Unofficial)/"
PATH_DATASETS <- "../datasets/"
PATH_DATA     <- "data/"
PATH_OUTPUT   <- "output/"

if (!dir.exists(PATH_DATA))   dir.create(PATH_DATA)
if (!dir.exists(PATH_OUTPUT)) dir.create(PATH_OUTPUT)

print("Setup complete ✅")
```

# =============================================================================
# 1. LOADING DATASETS
# =============================================================================

We work with **3 public and verifiable sources**:

| Dataset | Source | Description |
|---------|--------|-------------|
| FIFA WC 2026 Match Data | Kaggle (unofficial) | Fixtures, teams, venues and tournament groups |
| FIFA/Coca-Cola Rankings | FIFA official | Scores for 210 national teams (January 2026) |
| FIFA WC All Dataset | Gigasheet | World Cup historical data 1930–2022 |

> **Note:** Kaggle datasets are unofficial but built from FIFA's public information.
> Rankings are official and updated as of January 2026.

```{r 01 - Carga de datasets, include=TRUE}
print("=== LOADING DATASETS ===")
matches      <- read_csv(paste0(PATH_RAW, "matches.csv"),           show_col_types = FALSE)
teams        <- read_csv(paste0(PATH_RAW, "teams.csv"),             show_col_types = FALSE)
venues       <- read_csv(paste0(PATH_RAW, "host_cities.csv"),       show_col_types = FALSE)
stages       <- read_csv(paste0(PATH_RAW, "tournament_stages.csv"), show_col_types = FALSE)
rankings_raw <- read_csv(paste0(PATH_DATASETS, "fifa_rankings.csv"),show_col_types = FALSE)
hist         <- read_csv(paste0(PATH_DATASETS,
                 "FIFA World Cup 1930-2022 All Match Dataset.csv"),
                 show_col_types = FALSE)

cat("matches :", nrow(matches), "rows |", ncol(matches), "cols\n")
cat("teams   :", nrow(teams),   "rows |", ncol(teams),   "cols\n")
cat("venues  :", nrow(venues),  "rows |", ncol(venues),  "cols\n")
cat("stages  :", nrow(stages),  "rows |", ncol(stages),  "cols\n")
cat("rankings:", nrow(rankings_raw), "rows |", ncol(rankings_raw), "cols\n")
cat("hist    :", nrow(hist),    "rows |", ncol(hist),    "cols\n")
```

# =============================================================================
# 2. DATA VALIDATION
# =============================================================================

## Quality validation
Before any analysis, we validate data integrity using the `pointblank` package.
We check 4 key dimensions of the **Data Quality Framework**:

- **Completeness:** No null values in critical columns (IDs, dates)
- **Validity:** Positive IDs, ISO-formatted dates
- **Uniqueness:** No duplicate records
- **Consistency:** Logical ranges across all variables

Detailed HTML reports are saved to `output/` for auditing purposes.

## Cleaning and name standardization
A common challenge in data projects is that **different sources use different names**
for the same entity. For example, `fifa_rankings.csv` uses `"Korea Republic"` while
`teams.csv` uses `"South Korea"`.

Every correction is explicitly documented in the `case_when()` to keep the process
reproducible and auditable. We found **3 name inconsistencies** and **1 encoding issue**
(`Côte d'Ivoire`).

```{r 02 Validación con pointblank, include=TRUE}
al <- action_levels(warn_at = 0.05, stop_at = 0.20)

agent_matches <- create_agent(
  tbl = matches, tbl_name = "matches",
  label = "Data Validation World Cup 2026 - Matches", actions = al
) %>%
  col_vals_not_null(columns = vars(home_team_id, away_team_id, city_id, stage_id, kickoff_at)) %>%
  col_vals_gt(columns = vars(home_team_id, away_team_id, city_id), value = 0) %>%
  col_vals_regex(columns = vars(kickoff_at), regex = "^\\d{4}-\\d{2}-\\d{2}") %>%
  rows_distinct(columns = vars(id)) %>%
  interrogate()

cat("=== MATCHES REPORT ===\n")
print(agent_matches)

agent_teams <- create_agent(
  tbl = teams, tbl_name = "teams",
  label = "Data Validation World Cup 2026 - Teams", actions = al
) %>%
  col_vals_not_null(columns = everything()) %>%
  rows_distinct() %>%
  interrogate()

cat("\n=== TEAMS REPORT ===\n")
print(agent_teams)

agent_venues <- create_agent(
  tbl = venues, tbl_name = "venues",
  label = "Data Validation World Cup 2026 - Venues", actions = al
) %>%
  col_vals_not_null(columns = everything()) %>%
  rows_distinct() %>%
  interrogate()

cat("\n=== VENUES REPORT ===\n")
print(agent_venues)

get_agent_report(agent_matches) %>%
  export_report(filename = paste0(PATH_OUTPUT, "report_matches.html"))
get_agent_report(agent_teams) %>%
  export_report(filename = paste0(PATH_OUTPUT, "report_teams.html"))
get_agent_report(agent_venues) %>%
  export_report(filename = paste0(PATH_OUTPUT, "report_venues.html"))

print("Reports saved to output/ ✅")
```

# =============================================================================
# 3. CLEAN MERGE (Safe joins)
# =============================================================================

## Data integration (Merge)
We join all 4 datasets into a single dataframe (`data_full`) using `left_join()`
on numeric IDs. The result is a table of **104 matches** with all the information
needed for the analysis:

- Teams (name, group, code)
- Venue (stadium, city, country)
- FIFA Rankings (points, position)
- Derived variables (`rank_diff`, `points_diff`, `is_mexico`)

> `has_teams = TRUE` filters out fixtures that still have placeholder teams
> (e.g., "Winner UEFA Playoff D"), which exist at the start of the tournament.

```{r 03 - Limpieza y homologación de rankings, include=TRUE}
rankings_clean <- rankings_raw %>%
  clean_names() %>%
  filter(!is.na(rk)) %>%
  rename(rank = rk, team_name = team, points = total_points,
         prev_points = previous_points, change = x) %>%
  select(rank, team_name, points) %>%
  mutate(team_name = case_when(
    team_name == "Korea Republic"               ~ "South Korea",
    team_name == "Türkiye"                      ~ "Turkey",
    team_name == "Congo DR"                     ~ "DR Congo",
    str_detect(team_name, "Cura")               ~ "Curaçao",
    str_detect(team_name, "Ivoire|ivoire|C.te") ~ "Côte d'Ivoire",
    TRUE ~ team_name
  ))

cat("Clean rankings:", nrow(rankings_clean), "countries\n\n")
cat("🔍 Checking problematic teams:\n")
rankings_clean %>%
  filter(team_name %in% c("USA","Côte d'Ivoire","IR Iran","Curaçao","South Korea","Turkey")) %>%
  print()

write_csv(rankings_clean, paste0(PATH_DATA, "fifa_rankings_clean.csv"))
print("Saved: data/fifa_rankings_clean.csv ✅")
```

# =============================================================================
# 4. MAIN MERGE
# =============================================================================

```{r 04 - Merge principal, include=TRUE}
teams_clean <- teams %>%
  rename(team_id = 1, team_name = 2, team_code = 3,
         group = 4, is_placeholder = 5)

venues_clean <- venues %>%
  rename(city_id = 1, city_name = 2, country = 3,
         stadium = 4, region = 5, code = 6)

data_full <- matches %>%
  mutate(match_date = as.Date(kickoff_at)) %>%
  left_join(teams_clean %>% select(team_id, team_name, team_code, group, is_placeholder),
            by = c("home_team_id" = "team_id")) %>%
  rename(home_name = team_name, home_code = team_code,
         home_group = group, home_placeholder = is_placeholder) %>%
  left_join(teams_clean %>% select(team_id, team_name, team_code, group, is_placeholder),
            by = c("away_team_id" = "team_id")) %>%
  rename(away_name = team_name, away_code = team_code,
         away_group = group, away_placeholder = is_placeholder) %>%
  left_join(venues_clean %>% select(city_id, city_name, country, stadium),
            by = "city_id") %>%
  left_join(rankings_clean, by = c("home_name" = "team_name")) %>%
  rename(home_rank = rank, home_points = points) %>%
  left_join(rankings_clean, by = c("away_name" = "team_name")) %>%
  rename(away_rank = rank, away_points = points) %>%
  clean_names() %>%
  mutate(
    rank_diff   = coalesce(home_rank,   50L)   - coalesce(away_rank,   50L),
    points_diff = coalesce(home_points, 1400)  - coalesce(away_points, 1400),
    has_teams   = !home_placeholder & !away_placeholder,
    is_mexico   = country == "Mexico"
  )

cat("Total matches    :", nrow(data_full), "\n")
cat("With real teams  :", sum(data_full$has_teams, na.rm = TRUE), "\n")
cat("At Mexico venue  :", sum(data_full$is_mexico, na.rm = TRUE), "\n")

cat("\n🔍 Matches with match_date NA:", sum(is.na(data_full$match_date)), "\n")
data_full %>% filter(is.na(match_date)) %>% select(id, kickoff_at) %>% print()

write_csv(data_full, paste0(PATH_DATA, "mundial_2026_full.csv"), na = "")
write_rds(data_full, paste0(PATH_DATA, "mundial_2026_full.rds"))
print("Saved: data/mundial_2026_full.csv | data/mundial_2026_full.rds ✅")
```

# =============================================================================
# 5. RANKINGS COVERAGE CHECK
# =============================================================================

## Rankings coverage
We verify that every real tournament team has an assigned FIFA ranking.
100% coverage ensures the simulation model will not fall back to default values
(`1400 pts`) for any qualified team.

```{r 05 - Validar cobertura de rankings, include=TRUE}
teams_reales <- teams_clean %>% filter(is_placeholder == FALSE) %>% pull(team_name)
con_ranking  <- teams_reales[teams_reales %in% rankings_clean$team_name]
sin_ranking  <- teams_reales[!teams_reales %in% rankings_clean$team_name]

cat("Real tournament teams :", length(teams_reales), "\n")
cat("With FIFA ranking     :", length(con_ranking), "✅\n")
cat("Without ranking       :", length(sin_ranking), "⚠\n\n")

if (length(sin_ranking) > 0) {
  cat("Teams to add in case_when (chunk 03):\n")
  print(sin_ranking)
} else {
  print("100% coverage — all teams have a ranking")
}
```

# =============================================================================
# 6. MONTE CARLO SIMULATION SETUP
# =============================================================================

## Simulation model

### Model assumptions
The predictive model is based on the following assumptions:

1. **Team strength** is well approximated by FIFA points
2. **Win probability** follows a logistic function over the points difference
3. **Goals** follow a Poisson distribution with a mean proportional to the strength difference
4. **Mexico's home advantage:** +8% over the base probability (adjustable)
5. **In knockout rounds**, draws are resolved by penalties (50/50 probability)

### Known limitations
- Does not account for injuries, suspensions, or recent form
- Placeholder teams (yet-to-qualify via playoffs) are excluded until known
- The home boost is an estimate; it can be calibrated using historical World Cup data

```{r 06 - Simulación Monte Carlo (funciones), include=TRUE}
sim_match <- function(points_home, points_away, home_boost = 0) {
  diff   <- (points_home - points_away) / 600 + home_boost
  p_home <- plogis(diff + 0.1)
  p_away <- plogis(-diff)
  p_draw <- 1 - p_home - p_away
  probs  <- pmax(c(p_home, p_draw, p_away), 0.05)
  probs  <- probs / sum(probs)
  sample(c("home", "draw", "away"), size = 1, prob = probs)
}

sim_group <- function(data_full, teams_ranked, mexico_boost = 0.08) {
  results <- list()
  grupos  <- data_full %>%
    filter(has_teams == TRUE, !is.na(home_group)) %>%
    pull(home_group) %>% unique()

  for (g in grupos) {
    group_matches <- data_full %>% filter(home_group == g, has_teams == TRUE)
    group_teams   <- unique(c(group_matches$home_name, group_matches$away_name))
    standings     <- tibble(team = group_teams, pts = 0, gf = 0, ga = 0, played = 0)

    for (i in seq_len(nrow(group_matches))) {
      h     <- group_matches$home_name[i]
      a     <- group_matches$away_name[i]
      h_pts <- teams_ranked$points[teams_ranked$team == h]
      a_pts <- teams_ranked$points[teams_ranked$team == a]
      h_pts <- ifelse(length(h_pts) == 0 || is.na(h_pts), 1400, h_pts)
      a_pts <- ifelse(length(a_pts) == 0 || is.na(a_pts), 1400, a_pts)
      boost  <- ifelse(group_matches$is_mexico[i], mexico_boost, 0)
      result <- sim_match(h_pts, a_pts, home_boost = boost)
      lh <- exp(0.3 + (h_pts - a_pts) / 2000)
      la <- exp(0.3 - (h_pts - a_pts) / 2000)
      gh <- rpois(1, pmin(lh, 5))
      ga <- rpois(1, pmin(la, 5))
      if (result == "home") {
        standings$pts[standings$team == h] <- standings$pts[standings$team == h] + 3
      } else if (result == "away") {
        standings$pts[standings$team == a] <- standings$pts[standings$team == a] + 3
      } else {
        standings$pts[standings$team == h] <- standings$pts[standings$team == h] + 1
        standings$pts[standings$team == a] <- standings$pts[standings$team == a] + 1
      }
      standings$gf[standings$team == h] <- standings$gf[standings$team == h] + gh
      standings$ga[standings$team == h] <- standings$ga[standings$team == h] + ga
      standings$gf[standings$team == a] <- standings$gf[standings$team == a] + ga
      standings$ga[standings$team == a] <- standings$ga[standings$team == a] + gh
      standings$played[standings$team == h] <- standings$played[standings$team == h] + 1
      standings$played[standings$team == a] <- standings$played[standings$team == a] + 1
    }
    standings  <- standings %>%
      mutate(gd = gf - ga, group = g) %>%
      arrange(desc(pts), desc(gd), desc(gf))
    results[[g]] <- standings
  }

  all_standings <- bind_rows(results)
  top2 <- all_standings %>%
    group_by(group) %>% slice_head(n = 2) %>%
    mutate(classified_as = ifelse(row_number() == 1, "1st", "2nd")) %>% ungroup()
  terceros <- all_standings %>%
    group_by(group) %>% slice(3) %>% ungroup() %>%
    arrange(desc(pts), desc(gd), desc(gf)) %>%
    slice_head(n = 8) %>% mutate(classified_as = "3rd")
  bind_rows(top2, terceros) %>% select(team, pts, gd, gf, group, classified_as)
}

sim_knockout <- function(teams_vec, teams_ranked) {
  winners <- c()
  for (i in seq(1, length(teams_vec), by = 2)) {
    h     <- teams_vec[i];  a <- teams_vec[i + 1]
    h_pts <- teams_ranked$points[teams_ranked$team == h]
    a_pts <- teams_ranked$points[teams_ranked$team == a]
    h_pts <- ifelse(length(h_pts) == 0 || is.na(h_pts), 1400, h_pts)
    a_pts <- ifelse(length(a_pts) == 0 || is.na(a_pts), 1400, a_pts)
    result <- sim_match(h_pts, a_pts)
    winner <- if (result == "home") h else if (result == "away") a else sample(c(h, a), 1)
    winners <- c(winners, winner)
  }
  winners
}

run_tournament <- function(data_full, teams_ranked, mexico_boost = 0.08) {
  clasificados <- sim_group(data_full, teams_ranked, mexico_boost)
  equipos_kn   <- clasificados$team
  while (length(equipos_kn) < 32) equipos_kn <- c(equipos_kn, sample(clasificados$team, 1))
  equipos_kn <- equipos_kn[1:32]
  r32     <- sim_knockout(equipos_kn, teams_ranked)
  r16     <- sim_knockout(r32,        teams_ranked)
  qf      <- sim_knockout(r16,        teams_ranked)
  sf      <- sim_knockout(qf,         teams_ranked)
  campeon <- sim_knockout(sf,         teams_ranked)
  list(champion = campeon, finalist = sf, semi = qf, quarter = r16)
}

print("Simulation functions loaded (F1–F4) ✅")
```

# =============================================================================
# 7. RUNNING MONTE CARLO SIMULATION [10k]
# =============================================================================

```{r 07 - Simulación Monte Carlo (ejecución 10k), include=TRUE}
teams_ranked <- data_full %>%
  filter(has_teams == TRUE) %>%
  select(home_name, home_points) %>%
  rename(team = home_name, points = home_points) %>%
  bind_rows(
    data_full %>% filter(has_teams == TRUE) %>%
      select(away_name, away_points) %>%
      rename(team = away_name, points = away_points)
  ) %>%
  filter(!is.na(points)) %>%
  distinct(team, .keep_all = TRUE)

cat("Teams in simulation:", nrow(teams_ranked), "\n\n")

set.seed(2026)
N_SIMS    <- 10000
champions <- character(N_SIMS)
finalists <- character(N_SIMS * 2)

cat("🔄 Running", N_SIMS, "simulations...\n")
start_time <- Sys.time()

for (i in seq_len(N_SIMS)) {
  result       <- run_tournament(data_full, teams_ranked)
  champions[i] <- result$champion
  finalists[((i - 1) * 2 + 1):(i * 2)] <- result$finalist
  if (i %% 2000 == 0) cat(" →", i, "completed...\n")
}

elapsed <- round(as.numeric(Sys.time() - start_time, units = "secs"), 1)
cat("Completed in", elapsed, "seconds ✅\n")
```

# =============================================================================
# 8. RESULTS & EXPORT
# =============================================================================

## Monte Carlo simulation results
We ran **10,000 complete tournaments** with `set.seed(2026)` to guarantee
reproducibility. Results reflect statistical probabilities based exclusively
on current FIFA rankings and the tournament structure.

### Key observations
- **Spain, Argentina, and France** account for ~19% of championship probability,
consistent with their dominance in the FIFA rankings
- **Mexico** ranks **#11 for champion (3.63%)** and **#9 for reaching the final (7.39%)**,
boosted by its home-venue advantage
- The distribution is relatively flat (6.61% max), reflecting the new 48-team format
where tournament variance is higher than in previous editions

```{r 08 - Resultados y exportación, include=TRUE}
prob_campeon <- table(champions) %>%
  as_tibble() %>%
  rename(team = champions, n_wins = n) %>%
  mutate(prob_title = round(n_wins / N_SIMS * 100, 2)) %>%
  arrange(desc(prob_title))

prob_final <- table(finalists) %>%
  as_tibble() %>%
  filter(finalists != "") %>%
  rename(team = finalists, n_final = n) %>%
  mutate(prob_final = round(n_final / N_SIMS * 100, 2)) %>%
  arrange(desc(prob_final))

cat("🏆 TOP 15 - CHAMPION PROBABILITY:\n")
cat("══════════════════════════════════\n")
prob_campeon %>% slice_head(n = 15) %>% print(n = 15)

cat("\n🥈 TOP 10 - FINALIST PROBABILITY:\n")
cat("══════════════════════════════════\n")
prob_final %>% slice_head(n = 10) %>% print(n = 10)

cat("\n🇲🇽 MEXICO METRICS:\n")
cat("════════════════════\n")
mex_title <- prob_campeon %>% filter(team == "Mexico") %>% pull(prob_title)
mex_final <- prob_final   %>% filter(team == "Mexico") %>% pull(prob_final)
cat("Champion probability :", ifelse(length(mex_title) > 0, paste0(mex_title, "%"), "< 0.01%"), "\n")
cat("Finalist probability :", ifelse(length(mex_final) > 0, paste0(mex_final, "%"), "< 0.01%"), "\n")

write_csv(prob_campeon, paste0(PATH_OUTPUT, "sim_prob_campeon.csv"))
write_csv(prob_final,   paste0(PATH_OUTPUT, "sim_prob_final.csv"))
print("Exported: output/sim_prob_campeon.csv | output/sim_prob_final.csv ✅")
```

# =============================================================================
# 09 - VISUALIZATION: Championship Probabilities
# =============================================================================

## Visualizations
Charts are generated with `ggplot2` and exported at high resolution (300 dpi) to `output/`.
They are ready to be included in presentations, reports, or the GitHub README.

```{r 09 - Visualizacion Campeon, include=TRUE, fig.width=10, fig.height=8}
plot_campeon <- prob_campeon %>%
  slice_head(n = 15) %>%
  mutate(
    iso2 = tolower(countrycode(team, origin = 'country.name', destination = 'iso2c')),
    iso2 = ifelse(team == "England", "gb-eng", iso2)
  )

g_campeon <- ggplot(plot_campeon, aes(x = reorder(team, prob_title), y = prob_title)) +
  geom_segment(aes(xend = team, y = 0, yend = prob_title), color = "gray70", size = 1.2) +
  geom_flag(aes(y = prob_title + 0.2, image = iso2), size = 0.05) +
  geom_text(aes(y = prob_title + 0.8, label = paste0(prob_title, "%")),
            size = 4.5, fontface = "bold", color = "#F4D03F") +
  coord_flip() +
  scale_y_continuous(limits = c(0, max(plot_campeon$prob_title) + 1.5)) +
  labs(
    title    = "🏆 Probability of Winning the 2026 World Cup",
    subtitle = "Based on 10,000 Monte Carlo simulations using FIFA Rankings",
    x = "", y = "Probability (%)",
    caption  = "Data Analysis Portfolio"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.background  = element_rect(fill = "#1a1a2e", color = NA),
    panel.background = element_rect(fill = "#1a1a2e", color = NA),
    plot.title    = element_text(face = "bold", size = 18, margin = margin(b = 10), color = "white"),
    plot.subtitle = element_text(color = "gray60", size = 12, margin = margin(b = 20)),
    axis.text     = element_text(face = "bold", color = "#a0aec0", size = 12),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_line(color = "#2d3748", linewidth = 0.3)
  )

print(g_campeon)
ggsave(paste0(PATH_OUTPUT, "plot_campeon.png"), g_campeon, width = 10, height = 7, dpi = 300)
print("Chart saved: output/plot_campeon.png ✅")
```

# =============================================================================
# 10 - VISUALIZATION: Venue Distribution (Mexico focus)
# =============================================================================

```{r 10 - Visualizacion Sedes, include=TRUE, fig.width=10, fig.height=8}
plot_sedes <- data_full %>%
  count(stadium, city_name, country) %>%
  arrange(desc(n)) %>%
  mutate(
    sede_label = paste0(stadium, " (", city_name, ")"),
    color_pais = case_when(
      country == "Mexico" ~ "#006847",
      country == "USA"    ~ "#002868",
      country == "Canada" ~ "#FF0000",
      TRUE ~ "gray"
    )
  )

g_sedes <- ggplot(plot_sedes, aes(x = reorder(sede_label, n), y = n, fill = color_pais)) +
  geom_col(width = 0.7, alpha = 0.9) +
  geom_text(aes(label = n), hjust = -0.3, size = 4.5, fontface = "bold", color = "white") +
  coord_flip() +
  scale_fill_identity() +
  scale_y_continuous(limits = c(0, max(plot_sedes$n) + 1), breaks = seq(0, 10, by = 2)) +
  labs(
    title    = "🏟️ 2026 World Cup Matches by Stadium",
    subtitle = "Distribution of 104 matches across Mexico, United States, and Canada.",
    x = "", y = "Number of Matches",
    caption  = "Official FIFA data | Data Analysis Portfolio"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.background  = element_rect(fill = "#1a1a2e", color = NA),
    panel.background = element_rect(fill = "#1a1a2e", color = NA),
    plot.title    = element_text(face = "bold", size = 18, color = "white"),
    plot.subtitle = element_markdown(size = 12, color = "gray60"),
    axis.text     = element_text(size = 11, color = "#a0aec0"),
    panel.grid.major.y = element_blank(),
    legend.position = "none"
  )

print(g_sedes)
ggsave(paste0(PATH_OUTPUT, "plot_sedes.png"), g_sedes, width = 10, height = 8, dpi = 300)
print("Chart saved: output/plot_sedes.png ✅")
```

# =============================================================================
# 11: AUXILIARY CSVs FOR POWER BI
# =============================================================================

```{r 11 - CSV auxiliares Power BI, include=TRUE}
# Helper to write UTF-8 CSVs with BOM (ensures correct encoding in Power BI)
write_csv_utf8 <- function(x, path, ...) {
  tmp <- tempfile(fileext = ".csv")
  readr::write_csv(x, tmp, ...)
  raw <- readBin(tmp, what = "raw", n = file.info(tmp)$size)
  bom <- charToRaw("\ufeff")
  writeBin(c(bom, raw), path)
}

curiosidades <- tibble(
  categoria = c(
    "Países campeones","Finales más goleadoras","Goles promedio/edición",
    "Partidos más largos","Ediciones disputadas","Países participantes total",
    "Récord victorias consecutivas","Goles totales","% empates",
    "Victorias locales","Finales por penales","Remontadas épicas",
    "México en Mundiales","México títulos"
  ),
  valor = c(
    "8","1958 (9 goles)","2.8","120 min","22",
    "82 países","25 victorias (Brasil)","2,778","26%",
    "41%","5","12%","17/22","0"
  ),
  fuente = c(
    "Histórico","Suecia 1958","1930-2022","Prórrogas","1930-2022",
    "Histórico","1970","Histórico","Histórico","Histórico",
    "1994, 2006...","Estadística","Histórico","Histórico"
  )
)
write_csv_utf8(curiosidades, "../datasets/dim_curiosidades_mundial.csv", na = "")
cat("✅ dim_curiosidades_mundial.csv:", nrow(curiosidades), "rows\n")

# Load full historical dataset
hist_completo <- read_csv(
  "../datasets/FIFA World Cup 1930-2022 All Match Dataset.csv",
  show_col_types = FALSE
)

# Extract champions from traditional finals (20 editions)
campeones_finales <- hist_completo %>%
  filter(`Stage Name` == "final") %>%
  mutate(
    campeon = case_when(
      `Home Team Win` == 1 ~ `Home Team Name`,
      `Away Team Win` == 1 ~ `Away Team Name`,
      TRUE ~ NA_character_
    ),
    campeon = case_when(campeon == "West Germany" ~ "Germany", TRUE ~ campeon)
  ) %>%
  pull(campeon) %>% discard(is.na)

# Add Uruguay 1950 (round-robin edition, no traditional final)
campeones_completos    <- c(campeones_finales, "Uruguay")
dim_titulos_historicos <- tibble(team = campeones_completos) %>%
  count(team, name = "titulos_historicos") %>%
  arrange(dplyr::desc(titulos_historicos))

write_csv_utf8(dim_titulos_historicos, "../datasets/dim_titulos_historicos.csv", na = "")
cat("✅ dim_titulos_historicos.csv:", nrow(dim_titulos_historicos), "teams\n")
print(dim_titulos_historicos)

# ✅ FIX: object renamed from dim_jugadores_top → dim_leyendas (consistent)
dim_leyendas <- tibble(
  jugador = c(
    "Miroslav Klose","Ronaldo","Gerd Müller","Just Fontaine","Lionel Messi",
    "Peter Shilton","Fabien Barthez","Hugo Lloris","Manuel Neuer","Iker Casillas"
  ),
  tipo      = c(rep("Goleador", 5), rep("Portero", 5)),
  record    = c(16, 15, 14, 13, 13, 10, 9, 8, 7, 7),
  pais      = c("Germany","Brazil","Germany","France","Argentina",
                "England","France","France","Germany","Spain"),
  mundiales = c(4, 3, 2, 1, 5, 3, 3, 3, 2, 3)
) %>%
  arrange(tipo, desc(record)) %>%
  mutate(promedio_por_mundial = round(record / mundiales, 1))

write_csv_utf8(dim_leyendas, "../datasets/dim_leyendas.csv", na = "")
cat("✅ dim_leyendas.csv:", nrow(dim_leyendas), "players\n")
print(dim_leyendas)

# Convert historical data to long format: one row per team per match
partidos_largo <- bind_rows(
  hist_completo %>%
    transmute(fecha = lubridate::mdy(`Match Date`), team = `Home Team Name`, victoria = `Home Team Win`),
  hist_completo %>%
    transmute(fecha = lubridate::mdy(`Match Date`), team = `Away Team Name`, victoria = `Away Team Win`)
) %>%
  mutate(team = case_when(team == "West Germany" ~ "Germany", TRUE ~ team)) %>%
  arrange(team, fecha)

# Function to calculate maximum win streak using rle()
max_racha <- function(victorias) {
  if (all(victorias == 0)) return(0)
  r <- rle(victorias)
  max(r$lengths[r$values == 1], na.rm = TRUE)
}

dim_rachas_victorias <- partidos_largo %>%
  group_by(team) %>%
  summarise(
    partidos_totales        = n(),
    victorias_totales       = sum(victoria, na.rm = TRUE),
    record_victorias_consec = max_racha(victoria),
    .groups = "drop"
  ) %>%
  arrange(desc(record_victorias_consec))

write_csv_utf8(dim_rachas_victorias, "../datasets/dim_rachas_victorias.csv")
cat("✅ dim_rachas_victorias.csv:", nrow(dim_rachas_victorias), "teams\n")
print(dim_rachas_victorias, n = 15)
```

# =============================================================================
# 12. CONCLUSIONS
# =============================================================================

```{r conclusiones-inline, include=FALSE}
top1      <- prob_campeon %>% arrange(desc(prob_title)) %>% slice(1)
top2      <- prob_campeon %>% arrange(desc(prob_title)) %>% slice(2)
top3      <- prob_campeon %>% arrange(desc(prob_title)) %>% slice(3)
mex_pos   <- which(prob_campeon$team == "Mexico")
mex_title <- prob_campeon %>% filter(team == "Mexico") %>% pull(prob_title)
mex_final <- prob_final   %>% filter(team == "Mexico") %>% pull(prob_final)
```

The results of 10,000 simulations point to **`r top1$team`** as the clear favorite
with **`r top1$prob_title`%** championship probability, followed by **`r top2$team`**
(**`r top2$prob_title`%**) and **`r top3$team`** (**`r top3$prob_title`%**).

Three key findings:

1. **The ranking–probability correlation is robust.** The scatter confirms the model is
   coherent — higher FIFA points consistently translate into higher championship probability.

2. **The 48-team format opens up the tournament.** With a maximum individual probability
   of just `r top1$prob_title`%, no team is an overwhelming favorite; uncertainty is higher
   than in any previous edition.

3. **Mexico has a genuine shot as host.** With **`r mex_title`%** championship probability
   and **`r mex_final`%** probability of reaching the Final, Mexico ranks **#`r mex_pos`**
   globally — firmly within the group of teams with realistic chances. The +8% home boost
   reflects the local advantage documented in the sports modeling literature.

---

# 13. About this project {.unnumbered}

This project is part of my **data analysis portfolio**, developed from January
to March 2026. It demonstrates skills across:

| Area | Tools |
|------|-------|
| **Data engineering** | R · tidyverse · janitor · lubridate |
| **Data quality** | pointblank (Data Quality Framework) |
| **Applied statistics** | Logistic model · Monte Carlo simulation |
| **Visualization** | ggplot2 · ggimage · Power BI (5 pages) |
| **Reproducible reporting** | R Markdown → self-contained HTML |
| **Version control** | Git · GitHub |

> 📊 **Interactive dashboard** available in Power BI (5 pages with HTML Content lite)  
> 💻 **Source code** available on GitHub  
> 📧 **Contact:** [joel.giroud01@gmail.com](mailto:joel.giroud01@gmail.com) · [LinkedIn](https://linkedin.com/in/joel-pgiroud)

---

*Data: Official FIFA · Kaggle · Data Analysis Portfolio · 2026*
